---
title: Exploratory Data Analysis of Survey Data
author: Eric Leung
date: 2018-09-12
output:
  html_notebook:
    toc: true
    toc_float: true
---

# Overview

This report serves as an initial look at freeCodeCamp's 2018 New Coder Survey
data.

This data has not been pre-processed in any way. Thus, this report also serves
to explore necessary pre-processing to clean the data.

In exploring possible pre-processing steps, I will try to create thresholds and
rules on filtering out data that I find disingenuous for the sake of keeping
the integrity of the data high. So possible reasons for removing a row (i.e. a
unique survey submission from the final dataset) are

- Extremely high or low answers
- Submitting irrelevant answers

**Last Updated**: `r date()`


# Load Packages

```{r Load Packages}
# General help on manipulating data and visualize
library(tidyverse)
library(here)

# More specialized exploration tools
library(Hmisc)
```


# Read in Data

Let's read in the data.

```{r Read in Data}
# Read in data
raw_data_path <- here("raw-data", "2018-New-Coders-Survey.csv")
raw_data <- read_csv(raw_data_path)

# Explore dimensions
data_dim <- dim(raw_data)
```

Just a general check, there were **`r data_dim[1]` rows** and **`r data_dim[2]`
columns** in the raw data.

# Remove Basic Columns

Here, let's remove some columns we know will not be useful in the final dataset.

- Unique ID (unnecessary in the final dataset as we'll assume each row is
  independent individual)

This variable is encoded with the `#` symbol.

```{r}
# Remove unique ID column
raw_data <- raw_data %>% select(-"#")
```


# Explore Data

## Initial Overview

With all the data of interest at hand, let's do a very simple exploration of
the data using the `Hmisc` function `describe()`.

This function goes through each variable and gives a summary that is data-type
specific.

```{r Start to Explore Data}
# Basic overview of the data
Hmisc::describe(raw_data)
```

From this output, here are some notes about the data that will have to be
addressed in the pre-processing step.

- For lots of the data, it was encoded as a `1` or `0`, when it really should
  have been a boolean value `TRUE` or `FALSE`. Here are the columns that need
  this treatment.
    - "Are you already working as a software developer?"
    - "Is this your first software development job?"
    - "Are you willing to relocate for a job?"
    - "Have you attended a full-time coding bootcamp?"
    - "Have you finished yet?"
    - "Did you take a loan to pay for the bootcamp?"
    - "Based on your experience, would you recommend this bootcamp to your
      friends?"
    - "Are you an ethnic minority in your country?"
    - "Do you financially support any dependents?"
    - "Do you have children?"
    - "Do you financially support any elderly relatives or relatives with
      disabilities?"
    - "Do you have any debt?"
    - "Do you have a home mortgage?"
    - "Do you have student loan debt?"
    - "Do you consider yourself under-employed?"
    - "Have you served in your country's military before?"
    - "Do you receive disability benefits from your government?"
    - "Do you have high speed internet at your home?"
- The column on "Before you got this job, how many months did you spend looking
  for a job?" has some suspicious values e.g. 200 months = 16 years.
- Need to rename job preference to be the appropriate column label
- Need to re-code job preferences to be boolean, not just the string itself
- Need to explore and normalize (sanity check) the "Other" category for job
  preferences and group them accordingly
- There are some very high salary estimates for a first developer job that need
  further looking into
  encoding
- Search through the "Other" reasons on wanting to code and normalize them
  accordingly.
- The online resources need to be recoded into sensible column variables and
  their values need to be boolean as well
- The "Other" answers for online resources need to be checked and text
  normalized
- Coding resource also needs to be recoded with sensible column variables and
  values
- The "Other" answers for coding resources needs to be text normalized
- The coding podcasts and "Other" answers need to be normalized and recoded
  accordingly
- The YouTube channel and "Other" answers need to be normalized and recoded
  accordingly
- The number of hours you spend learning each week needs to have a threshold as
  people put they have spent all 168 hours in a week studying. A sensible
  number *might* be around 126 hours (given 6 hours of sleep everyday)
- The number of months spent programming needs to be looked at and have a
  threshold
  because some people claim to have programmed for 55 years.
- Money spent on coding needs to have a threshold and checked because of
  possible extreme values
- Check age distribution for outliers (e.g. someone is 250 years old)
- Check values for "Other" genders for irrelevant answers and text normalize if
  necessary
- Find threshold for number of children (e.g. someone claims to have 100
  children) so a sensible number may be up to 7
- Check debt owed for a home mortgage, there are some extreme values
- Check debt owed for student debt, there are some extreme values
- Check values for "Other" in current employment situation, there may be some
  irrelevant values
- Check money made last year, there are some extreme values
- Check average time of survey completion for curiosity


## Column Specific Checks

After getting a brief overview of the data, let's take a look at some of the
individual columns that were flagged in the previous section.

### Specify Free Text Variables

Here, let us specify the variables that had free text. This will allow us to
identify rows/entries that had low effort or are irrelevant. These are
specifically the non-numeric variables to make it easier to spot.

```{r}
free_text_cols <- c(
  "Other",
  "Other_1",
  "Other_2",
  "Other_3",
  "Other_4",
  "Other_5",
  "Which one?",
  "Other_6",
  "Other_7"
)
```


### Number of months spent looking for a job

```{r}
month_quantiles <- raw_data %>%
    select(`Before you got this job, how many months did you spend looking for a job?`) %>%
    pull() %>%
    quantile(probs = c(0.95, 0.99, 0.999, 0.9999, 1), na.rm = TRUE)
month_quantiles
```

Looking at the above distribution, it looks like we may have to remove
individuals (or at least investigate individuals) with months greater than
`r month_quantiles[2]` months.


### Check expected salary amounts

The question asked was

> About how much money do you expect to earn per year at your first developer
> job (in US Dollars)?

As noted above, the extremes of estimated first developer job salary don't
seem like they were from well-intentioned individuals. So here, let's take a
look at the distribution more closely.

```{r}
raw_data %>%
  rename(money_expect = `About how much money do you expect to earn per year at your first developer job (in US Dollars)?`) %>%
  select(money_expect) %>%
  filter(!is.na(money_expect)) %>%
  qplot(bins = 30,
        xlab = "Expected Salary",
        ylab = "Count",
        main = "Distribution of Expected Job Salary")
```

We see there is a number of individuals who are creating a large skew in the
expected salary from a job.

```{r}
raw_data %>%
  rename(money_expect = `About how much money do you expect to earn per year at your first developer job (in US Dollars)?`) %>%
  filter(money_expect > 250000) %>%
  select(money_expect, free_text_cols) %>%
  arrange(desc(money_expect)) %>%
  data.frame() %>%
  describe()
```

There doesn't appear to be any irrelevant comments from these variables, so we
can probably just leave this variable alone in cleaning.

### Read some of the reasons why people want to learn how to code

The "Other" category for the reason for wanting to code can be very
interesting. Here, let's explore some of them and see if we have to worry about
some of them being outliers.

```{r}

```


### Check number of hours spent learning

```{r}

```


### Check number of months programming

```{r}

```

### Check month spent on programming outside of university

```{r}

```


### Check distribution of ages

```{r}

```

### Check other genders

```{r}

```


### Check number of children

```{r}

```


### Check debt owed for home mortgage

```{r}

```


### Check debt owed for student debt

```{r}

```


### Check "Other" answer for employment status

```{r}

```


### Check how much money made last year

```{r}

```


### Check distribution of survey completion times

```{r}

```


## Check Other Columns

### Resources

```{r}
raw_data %>%
  select(Other_2) %>%
  filter(!is.na(Other_2)) %>%
  count(Other_2) %>%
  arrange()
```



# Check Outlier Rich Columns

There are some columns that had values quite extreme from everyone else. Here,
let's clump all of those columns together to see if there are any patterns.

```{r}

```


# Check Empty Rows

Some entries to the survey may have not be serious. In other words, these
individuals may have filled most of the survey but then left the rest of it
blank. Here, I test my hypothesis that there are such people.

First we'll need to calculate how sparse each of the rows are. This should be a
simple count the number of `NA` values for each column, with respect to the
total number of columns.

```{r Check Sparse Rows}
# Calculate sparsity of rows
num_cols <- ncol(raw_data)
sparsity_perc <- raw_data %>%
  is.na() %>%
  rowSums() / num_cols
sparsity_perc %>% summary()
```

Looking at these statistics, we can see that there are some rows with the
majority of the columns are `NA` values. So we can probably remove these
safely.

```{r}
sparsity_perc %>% stem()
```

As a quick-and-dirty visualization, it looks like the majority of the entries
are `NA` values, specifically around 70%. This could be because of lots of the
multiple answer questions.

But looking at this distribution, even the other extreme is suspicious, namely
the very complete entries, where only 13% of the values are empty.

# Distribution of Completion Times

It might be interesting to look at the distribution of completion times. This
can be valuable because individuals who finished really quickly, may have put
down irrelevant answers that wouldn't be useful.

```{r}
completion_time <- raw_data %>%
  select(`Submit Date (UTC)`, `Start Date (UTC)`) %>%
  mutate(total_time_sec = `Submit Date (UTC)` - `Start Date (UTC)`) %>%
  mutate(total_time_min = total_time_sec / 60) %>%
  mutate(total_time_min = as.numeric(total_time_min))
completion_time %>%
  pull(total_time_min) %>%
  stem()
```

There is just a few people who took a long time with this data, namely one
individual who took
`r (max_complete_time <- completion_time %>% select(total_time_min) %>% max)`
minutes to finish, which equates to about
`r max_complete_time / 60` hours to finish.

There are also individuals who wanted to get through the survey as fast as
possible.

```{r}
completion_time %>%
  filter(total_time_min < 1) %>%
  arrange(total_time_min) %>%
  as.data.frame() %>%
  head()
```

```{r}
completion_time %>%
  filter(total_time_min > 100) %>%
  arrange(total_time_min) %>%
  as.data.frame() %>%
  head()
```

# Summary of Pre-Processing Data

Here is a summary of what data pre-processing needs to be done.

- Remove ages greater than 116 (and be skeptical of those around that age)
